Covariant Phase Space via MultiSymplectic Geometry

### 2017-05-26 cpsviams.tex
Almost Finished!

\documentclass[a4paper,12pt]{scrartcl}  %Per La Stampa
\usepackage[a4paper, margin=2cm]{geometry}

@@

\begin{abstract}
La costruzione del covariant phase space con la cognizione del formalismo simplettico come descritta nell'articolo di forger romero.

@@

\section{Covariant phase space approach}
The gist of the transition from the \emph{multi-symplectic approach} to the \emph{Covariant Functional Approach} - based on the \emph{Covariant Phase Space} - is basically to perform a change of perspective.\\
We have to move our attention from the bundles to the cross-sections on the bundles.

@@

	\begin{definition}[Space of kinematics (off-shell) configurations]\label{Def:ConfSpace}
		Is the, generally non-linear, infinte-dimensional manifold :
		\begin{displaymath}
			\Conf \coloneqq \Gamma^\infty(F,M)
		\end{displaymath}
	\end{definition}

	\begin{definition}[Space of Dynamics (on-shell) configurations]\label{Def:SolSpace}
		Is the subset of $\Conf$:
		\begin{displaymath}
			\Sol \coloneqq \ker(D_\cdot) \subset \Conf
		\end{displaymath}
		containing all the smooth solutions of the motion equations corresponding to the  dynamical operator $D_\cdot$.
	\end{definition}

@@

Again, we would like to employ a collective notation to denote the motion equation operators $D_\Lagrangian$ and $D_\Hamiltonian$:
\begin{notation}

	Since operators $D$ when applied to $\phi \in \Conf$ yields a twisted 1-form, 
	we denote them collectively as $D[\phi] \in \Gamma(\phi^\ast(V^\TwistedLinear F))$.
	When $D$ is evaluated on a vector field yields
	\begin{displaymath}
		\left\langle D[\phi] \right\vert \left. V \right\rangle =
		\phi^\ast \left( V \lrcorner \omega \right) 
		\begin{cases}
			(j^1 \varphi)^\ast \left( j^1V \lrcorner \omega_\Lagrangian \right) \\
			(\varphi, \pi)^\ast \left( V \lrcorner \omega_\Hamiltonian \right)
		\end{cases}              
        \in \Omega^m(M)
	\end{displaymath}

	where has been employed the following correspondences:
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			 & Lagrangian picture & Hamiltonian picture \\
			\hline
			$\phi \in \Conf$		&	$j^1 \varphi$		&	$(\varphi, \pi)$	\\
			$V \in \Gamma(\phi^\ast V F )$ & $j^1 V$ & $V$ \\
			$\omega$ & $\omega_\Lagrangian$ & $\omega_\Hamiltonian$ \\
			\hline
		\end{tabular}
	\end{center}

	Noticing the recurring scheme we will make use of the following notation:
	\begin{displaymath}
		W = V^\TwistedLinear F \; ; \; V = \Gamma^\infty(\phi^\ast W)
	\end{displaymath}
	So,
	\begin{displaymath}
		D[\phi] \in V \quad \forall \phi \in \Conf
	\end{displaymath}
\end{notation}

@@

\section{Jacobi Operator}\label{Sec:JacobiOperator}
Intuitively, the Jacobi equation are the formal linearizazion of the motion equation around a solution $\phi$.\\
Precisely this can be encoded in the following operator:

	\begin{definition}[Jacobi operator pertaining to $D_\cdot$ around a solution $\phi \in \Sol$]\label{Def:JacobiOp}
		\begin{displaymath}
			J(\phi) \; : \; T_\phi \Conf \rightarrow	\phi^\ast(V^\TwistedLinear F) 
			\quad s.t. \quad  \left(J(\phi)\right) [ \delta\phi ]
			= \sigma \left( \left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \right)
		\end{displaymath}
	\end{definition}

\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/Jacobi.png}
\caption{\footnotesize \label{fig:jacobi}$ \left(J(\phi)\right) [ \delta\phi ]$ and $D$ target on the same space.}
\end{figure}
where:
\begin{itemize}
	\item $\phi_\lambda$ is an arbitrary variation of $\phi_0 = \phi$ generated by $\delta\phi$.
	\item exactly as noted for the tangent vector of $\Conf$, we have for any $x\in M$:
		\begin{displaymath}
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0}(x) \in V_M W
			\quad \Rightarrow \quad
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \in D[\phi]^\ast \left( V_M \left( V^\TwistedLinear \right)\right)
		\end{displaymath}
		because $D_\cdot[\phi_\lambda]$ can be regarded as a section of $V^\TwistedLinear F$ over $M$, 
		thus the latter definition does not depend from the particular choice of variation $\phi_\lambda$ 
		but only from its infinitesimal generator $\delta\phi$.
	\item $\sigma$ is the canonical bundle-morphism
		\begin{displaymath}
			\sigma:  0^\ast \left( V_M W \right)  \; \rightarrow \;  0^\ast \left(V_E W \right) \simeq W
		\end{displaymath}
		 where $V_M W$ and $V_F W$ are the vertical sub-bundle of $T W$ w.r.t. $M$ and $F$ respectively 
		 and $0$ is the zero-section of $W$ over $F$.
\end{itemize}

The operator is well defined since
\begin{displaymath}
	\phi \in \Sol \quad \Rightarrow \quad D[\phi] \subset 0(E)
\end{displaymath}

\begin{remark}
	The Jacobi operator $\left(J(\phi) \right) [\delta \phi]$ around a solution $\phi$ when acting on $\delta \phi \in T_\phi \Conf$ yields a twisted form that can be evaluated on a $V \in \Gamma\left(\phi^\ast(V^\TwistedLinear F )\right)$
\end{remark}

\subsection{Construction of $\sigma$}
Consider the two stacked bundle, $ W \rightarrow E$ vector bundle, $E \rightarrow M$ smooth bundle.

@@

We have the following short exact sequence of bundle morphism over $W$
\begin{figure}
  \centering
  \includestandalone[mode=buildnew]{sigma_exact}
\end{figure}
because

\begin{displaymath}
	\ker(i) = 0 (E) = i' \left( 0 \right)
\end{displaymath}
\begin{displaymath}
	\ker\left( T \pi_{E,W} \right) = V_E W = i\left( V_E W \right)
\end{displaymath}
\begin{displaymath}
	\ker\left(T \pi_{M,E}\right) = V_M E = T \pi_{E,W} \left( V_M W \right)
\end{displaymath}

@@

\newpage
\section{Covariant approach: pre-symplectic form}
\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/symform.png}
  \caption{(temp)}
\end{figure}

@@

\begin{proposition}\label{Prop:CurrentClosed}
	When restricted to $\textrm{T}_\phi \Sol$, the symplectic current is a closed form. That is:
	\begin{displaymath}
		\diff \Upsilon_\phi \left( \delta \phi_1 , \delta \phi_2 \right) = 0 \qquad \delta \phi_1 , \delta \phi_2 \in T_\phi \Sol
	\end{displaymath}
\end{proposition}
\begin{proof}
	Working in the collective notation introduced above, considering $\phi \in \Sol$ and $V_i \in T_\phi \Sol$  we have:
	\begin{align*}
		\diff \Upsilon_\phi \left( V_1 , V_2 \right) &= 
		\phi^\ast \diff \left[ i_{V_2} i_{V_1} \omega \right] = 
		\phi^\ast \left( - i_{V_2} \diff ( i_{V_1} \omega )  + \Lie_{V_2 } ( i_{V_1} \omega ) \right) =	\\
		&= \phi^\ast \left( 
			i_{V_2}i_{V_1} \diff \omega - 
			\Lie_{V_1 } ( i_{V_2} \omega ) +
			\Lie_{V_2 } ( i_{V_1} \omega ) +
			i_{[V_1,V_2]} \omega
		\right)
	\end{align*}
	using the Cartan's magic formula twice and the commutation rule between Lie derivative and contraction operator $i_\cdot$.\\
	The first term vanishes because $\omega= \diff \theta$ is exact by construction.\\
	The second and the third term vanish because
	\begin{displaymath}
		\left\langle \left( J(\phi) \right)[V_i] \right\vert \left. V_j \right\rangle = 
		\phi^\ast \Lie_{V_i} \left( i_{V_j} \omega \right)
	\end{displaymath}
	and $V_i$ are solution of the Jacobi equation
\end{proof}

@@

\begin{wrapfigure}{L}{0.41\textwidth}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/sigma.png}
  %\caption{.}
\end{wrapfigure}

@@

\bibitem{Gimmsy}
  Leslie Lamport,
  \emph{\LaTeX: a document preparation system},
  Addison Wesley, Massachusetts,
  2nd edition,
  1994.

>>>>

\documentclass[a4paper,12pt,fleqn]{scrartcl}  %Per La Stampa
\usepackage[a4paper, margin=2cm]{geometry}

@@

\begin{abstract}
In questo talk viene dettagliata la costruzione del covariant phase space partendo dal formalismo multisimplettico come descritto nell'artticolo di forger romero.

La costruzione del covariant phase space con la cognizione del formalismo simplettico come descritta nell'articolo di forger romero.

@@

\clearpage \newpage
\section{Covariant phase space approach}
The gist of the transition from the \emph{multi-symplectic approach} to the \emph{Covariant Functional Approach} - based on the \emph{Covariant Phase Space} - is basically to perform a change of perspective.\\
We have to move our attention from the bundles to the cross-sections on the bundles.

@@

	\begin{definition}[Space of kinematics (off-shell) configurations]\label{Def:ConfSpace}
		Is the, generally non-linear, infinte-dimensional manifold :
		\begin{displaymath}
			\Conf \coloneqq \Gamma^\infty(F,M)
		\end{displaymath}
	\end{definition}
%
	\begin{remark}
		Since the jet functor $j^1$ acts injectively on sections, we can equivalently consider any other holonomic jet sections
		\begin{displaymath}
			\Conf \simeq j^k \Gamma^\infty(F)
		\end{displaymath}
	\end{remark}	
%
	\begin{definition}[Space of Dynamics (on-shell) configurations]\label{Def:SolSpace}
		Is the subset of $\Conf$:
		\begin{displaymath}
			\Sol \coloneqq \ker(D_\cdot) \subset \Conf
		\end{displaymath}
		containing all the smooth solutions of the motion equations corresponding to the  dynamical operator $D_\Lagrangian$ or $D_\Hamiltonian$ depending on the framework considered.
	\end{definition}

@@

Again, we would like to employ a collective notation to denote the motion equation operators $D_\Lagrangian$ and $D_\Hamiltonian$:
\begin{notation}
	Since $D_\Lagrangian$ and $D_\Hamiltonian$ depend on different order jet bundle over $F$, can not be introduced directly a common notation on the multi-symplectic ground but we have to pass to the functional framework:
\begin{displaymath}
	D[\; ] = \; \Gamma^\infty(F) \ni \phi \quad \mapsto \quad D[ \phi ] \in \Gamma^\infty \left( \phi^\ast \left(V^\TwistedLinear F \right)\right)
\end{displaymath}	
in other words, $D[\phi](x) \in V^\TwistedLinear_{\phi(x)}F$ is a twisted linear one-form.
//
	The evaluation of operator $D$, when applied to $\phi$, on a test vector field $V \in \Gamma(\phi^\ast(VF))$, results:
	\begin{displaymath}
		\left\langle D[\phi] \right\vert \left. V \right\rangle =
		\phi^\ast \left( i_V \omega \right) 
		\begin{cases}
			(j^1 \varphi)^\ast \left( i_{j^1V} \: \omega_\Lagrangian \right) \\
			(\varphi, \pi)^\ast \left( i_V \: \omega_\Hamiltonian \right)
		\end{cases}              
        \in \Omega^m(M)
	\end{displaymath}

	where has been employed the following correspondences:
	\begin{center}
		\begin{tabular}{|c|c|c|}
			\hline
			 & Lagrangian picture & Hamiltonian picture \\
			\hline
			$\phi \in \Conf$		&	$j^1 \varphi \; , \quad \varphi \in \Gamma^\infty(E)$		&	$(\varphi, \pi)$	\\
			$V \in \Gamma(\phi^\ast V F )$ & $j^1 V$ & $V$ \\
			$\omega$ & $\omega_\Lagrangian$ & $\omega_\Hamiltonian$ \\
			\hline
		\end{tabular}
	\end{center}

	Noticing the recurring scheme we will also denote $
		W = V^\TwistedLinear F $
\end{notation}

@@

\clearpage \newpage
\section{Jacobi Operator}\label{Sec:JacobiOperator}
Intuitively, the Jacobi equation are the formal linearizazion of the motion equation around a solution $\phi$.\\
Precisely this can be encoded in the following operator:

	\begin{definition}[Jacobi operator pertaining to $D$ around a solution $\phi \in \Sol$]\label{Def:JacobiOp} is the bundle map over $F$:
		\begin{displaymath}
			J(\phi) \; : \; \phi^\ast(VF) \rightarrow	\phi^\ast(V^\TwistedLinear F) 
			\quad s.t. \quad  \left(J(\phi)\right) V
			= \sigma \left( \left.\dfrac{\diff}{\diff \lambda} D [\phi_\lambda](x) \right\rvert_{\lambda=0} \right)
		\end{displaymath}
		where $V\in (\phi^\ast(VF))_{\phi(x)} = V_{\phi(x)}F$ and $\phi_\lambda$ is a regular variation generated by a $\delta\phi$ such that $\delta\phi_x = V$ and
		$\sigma$ is the canonical bundle-morphism
		\begin{displaymath}
			\sigma:  0^\ast \left( V_M W \right)  \; \rightarrow \;  0^\ast \left(V_E W \right) \simeq W
		\end{displaymath}
		 where $V_M W$ and $V_F W$ are the vertical sub-bundle of $T W$ w.r.t. $M$ and $F$ respectively 
		 and $0$ is the zero-section of $W$ over $F$.
	\end{definition}
	As a bundle-map, $J$ can be naturally interpreted functionally as
	\begin{displaymath}
		\left(J(\phi)\right) [\: ] : 
		\Gamma^\infty\left(\phi^\ast(VF)\right) \rightarrow	
		\Gamma^\infty \left( 
		\phi^\ast(V^\TwistedLinear F)
		\right)
	\end{displaymath}
%
\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/Jacobi.png}
\caption{\footnotesize \label{fig:jacobi}$ \left(J(\phi)\right) [ \delta\phi ]$ and $D$ target on the same space.}
\end{figure}
%
Since $D[\phi_\lambda](x)$ is a smooth curve on $ V^\TwistedLinear F $ that is vertical w.r.t. $M$, the derivative involved has to be intended as the tangent velocity vector, thus $\dfrac{\diff}{\diff \lambda} D [\phi_\lambda]$ is the vertical vector field associated to the infinitesimal variation.
Exactly as noted for the tangent vector of $\Conf$, we have for any $x\in M$:
		\begin{displaymath}
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0}(x) \in V_M W
			\quad \Rightarrow \quad
			\left.\dfrac{\diff}{\diff \lambda} D_\cdot [\phi_\lambda]\right\rvert_{\lambda=0} \in D[\phi]^\ast \left( V_M \left( V^\TwistedLinear \right)\right)
		\end{displaymath}
		because $D_\cdot[\phi_\lambda]$ can be regarded as a section of $V^\TwistedLinear F$ over $M$, 
		thus the latter definition does not depend from the particular choice of variation $\phi_\lambda$ 
		but only from its infinitesimal generator $\delta\phi$.
%
The operator is well defined since
\begin{displaymath}
	\phi \in \Sol \quad \Rightarrow \quad D[\phi] \subset 0(E)
\end{displaymath}

\begin{proposition}\label{Prop:JacobieLie}
	The Jacobi operator $\left(J(\phi) \right) [\delta \phi]$
	around a solution $\phi$ when acting on an infinitesimal variation $\delta \phi \in T_\phi \Conf$
	yields a twisted form that can be evaluated on a $V \in \Gamma\left(V F \right)$
	such that:
	\begin{displaymath}	
		\left\langle \left(J(\phi) \right) [\delta \phi] \right\vert
		\left. V \right\rangle =
		\phi^\ast \; \Lie_{\delta \phi} \left( i_V \omega \right)
	\end{displaymath}
\end{proposition}
\begin{proof}
The explicit action of the Jacobi Operator can be reconstructed as follows:
\begin{displaymath}
	D[\phi_\cdot](\cdot): (\lambda, x^\mu) \mapsto
	\left( x^\mu, \phi_\lambda^A (x), \hat{D}[\phi_\lambda]_A(x) \right)
	\in V^\TwistedLinear F 
\end{displaymath}
%
\begin{align*}
	\frac{\partial}{\partial \lambda} \eval{D[\phi_\lambda](\cdot)}_{\lambda=0}:
	x \mapsto
	\biggr(& x^\mu, \phi_0^A(x), \hat{D}[\phi_0]_A(x),\\
	&	
	\eval{\dfrac{\partial \phi^A(x)}{\partial \lambda}}_{\lambda=0}
	 \dfrac{\partial}{\partial q^A} + 
	\eval{\dfrac{\partial \hat{D}[\phi_\lambda]_A(x)}{\partial \lambda}}_{\lambda=0}
	 \dfrac{\partial}{\partial q_A}	  \biggr)
	 \in V_M(V^\TwistedLinear F)
\end{align*}
\begin{align*}
	\left(J(\phi)\right) [\: ] (\cdot) =
	\sigma \frac{\partial}{\partial \lambda} \eval{D[\phi_\lambda](\cdot)}_{\lambda=0}:
	x \mapsto &
	\left( x^\mu, \phi_0^A(x), 0,
	\eval{\dfrac{\partial \hat{D}[\phi_\lambda]_A(x)}{\partial \lambda}}_{\lambda=0}
	 \dfrac{\partial}{\partial q_A}	  \right)
	 \in V_F(V^\TwistedLinear F) \\
	 & \qquad \qquad \qquad \vert\wr \\
	 &
		\left( x^\mu, \phi_0^A(x),
			\eval{\dfrac{\partial \hat{D}[\phi_\lambda]_A(x)}{\partial \lambda}}_{\lambda=0} \right)
	 		\in V^\TwistedLinear F 
\end{align*}
	Therefore for all $x\in M$:
	\begin{align*}
		\left\langle \left(J(\phi) \right) [\delta \phi] \right\vert
		\left. V \right\rangle_x &=
		\left\langle \left(J(\phi) \right) [\delta \phi] (x) \right\vert
		\left. V_x \right\rangle =
		\left\langle \left(J(\phi) \right) [\delta \phi] (x)
		 \right\vert
		\left. V_x \right\rangle =\\
		&=
		\left\langle\left.		
		\eval{\dfrac{\partial \hat{D}[\phi_\lambda]_A(x)}{\partial \lambda}}_{\lambda=0}
		\right\vert
		 V_x \right\rangle =
		 \eval{
		 	\dfrac{\partial}{\partial \lambda}
		\left\langle\left.			 	
		 	\hat{D}[\phi_\lambda]
		 			\right\vert
		 V \right\rangle _x
		 }_{\lambda=0}
	\end{align*}
	from the linearity of the duality evaluation.\\
	Plugging in the definition of motion operator we obtain 
	\begin{displaymath}
				\left\langle \left(J(\phi) \right) [\delta \phi] \right\vert
		\left. V \right\rangle =
		 \eval{
		 	\dfrac{\partial}{\partial \lambda}
				\left(
					\phi^\ast_\lambda \left( i_V \omega \right) \right)
		 }_{\lambda=0}	=
		 \phi_0^\ast
		 \eval{
				\dfrac{\partial}{\partial \lambda}
				\left(
			 		\left(\phi_\lambda \cdot \phi_0^{-1}\right)^\ast
					\left( i_V \omega \right) \right)
		 }_{\lambda=0}	=
		 		\phi_0^\ast \; \Lie_{\delta \phi} \left( i_V \omega \right)
	\end{displaymath}
	recognizing that $\left(\phi_\lambda \cdot \phi_0^{-1}\right)$ is the flux generated by $\delta \phi$.
\end{proof}

\clearpage \newpage
\subsection{Construction of $\sigma$}
Consider the two stacked bundle, $ W \rightarrow E$ vector bundle, $E \rightarrow M$ smooth bundle.

@@

We have the following short exact sequence of bundle morphism over $W$:
\begin{figure}
  \centering
  \includestandalone[mode=buildnew]{sigma_exact}
\end{figure}
since
%
\begin{displaymath}
	\ker(i) = 0 (E) = i' \left( 0 \right)
\end{displaymath}
\begin{displaymath}
	\ker\left( T \pi_{E,W} \right) = V_E W = i\left( V_E W \right)
\end{displaymath}
\begin{displaymath}
	\ker\left(T \pi_{M,E}\right) = V_M E = T \pi_{E,W} \left( V_M W \right)
\end{displaymath}

@@

\clearpage \newpage
\section{Covariant approach: pre-symplectic form}
\begin{figure}
\centering
\includegraphics[width=0.40\textwidth]{Pictures/symform.png}
  \caption{(temp)}
\end{figure}

@@

\begin{proposition}\label{Prop:CurrentClosed}
	When restricted to $\textrm{T}_\phi \Sol$, the symplectic current is a closed form. That is:
	\begin{displaymath}
		\diff \Upsilon_\phi \left( \delta \phi_1 , \delta \phi_2 \right) = 0 \qquad \delta \phi_1 , \delta \phi_2 \in T_\phi \Sol
	\end{displaymath}
\end{proposition}
\begin{proof}
	Working in the collective notation introduced above, considering $\phi \in \Sol$ and $V_i \in T_\phi \Sol$  we have:
	\begin{align*}
		\diff \Upsilon_\phi \left( V_1 , V_2 \right) &= 
		\phi^\ast \diff \left[ i_{V_2} i_{V_1} \omega \right] = 
		\phi^\ast \left( - i_{V_2} \diff ( i_{V_1} \omega )  + \Lie_{V_2 } ( i_{V_1} \omega ) \right) =	\\
		&= \phi^\ast \left( 
			i_{V_2}i_{V_1} \diff \omega - 
			\Lie_{V_1 } ( i_{V_2} \omega ) +
			\Lie_{V_2 } ( i_{V_1} \omega ) +
			i_{[V_1,V_2]} \omega
		\right)
	\end{align*}
	using the Cartan's magic formula twice and the commutation rule between Lie derivative and contraction operator $i_\cdot$.\\
	The first term vanishes because $\omega= - \diff \theta$ is exact by construction.\\
	The second and the third term vanish because, from proposition \ref{Prop:JacobieLie},
	\begin{displaymath}
		\left\langle \left( J(\phi) \right)[V_i] \right\vert \left. V_j \right\rangle = 
		\phi^\ast \Lie_{V_i} \left( i_{V_j} \omega \right)
	\end{displaymath}
	and $V_i$ are solution of the Jacobi equation.\\
	The last term vanishes since the lie bracket of two vertical vector field along $\phi$ is equal to the brackets of two vertical field over $F$ constant on each fiber.
\end{proof}

@@

@@

\bibitem{Gimmsy}
	Gotay, M. J., Isenberg, J., Marsden, J. E., Montgomery,	R. (1998). 
	Momentum Maps and Classical Relativistic
	Fields. Part I: Covariant Field Theory, (August 2004), 68. 
	Retrieved from http://arxiv.org/abs/physics/9801019

\bibitem{primer}
Benini, M., Dappiaggi, C., Hack, T.-P. (2013). Quantum Field Theory on Curved Backgrounds — a Primer. International Journal of Modern Physics A, 28(17), 1330023. https://doi.org/10.1142/S0217751X13300238

\bibitem{forgeromero}
Forger, M.,  Romero, S. V. (2005). Covariant poisson brackets in geometric field theory. Communications in Mathematical Physics, 256(2), 375–410. https://doi.org/10.1007/s00220-005-1287-8
